Inferring stellar rotation periods using Gaussian processes

### 2017-01-27 documents/ms.tex
adding figures
minor edits

In order to recover the rotation period of a simulated light curve using
Gaussian processes, we sample the following posterior PDF:
\begin{equation}
p({\bm \theta}\,|\,y) \propto \mathcal L(y\,|\,{\bm \theta}) p({\bm \theta}),
\end{equation}
\label{eq:posterior}
where $y$ are the light curve data, $\bm \theta$ are the hyperparameters
of the kernel described in equation \ref{eq:QP}, $\mathcal L$ is the
GP likelihood function, and $p({\bm \theta})$ is the prior on the
hyperparameters.  The GP likelihood is similar to the simple Gaussian likelihood
function used for optimisation problems where the uncertainties are
Gaussian and uncorrelated. The latter can be written
\begin{equation}
\ln \mathcal{L} = -\frac{1}{2}\sum_{n=1}^N\frac{(y_n-\mu)^2}{\sigma_n^2}
    - \frac{N}{2}\ln(2\pi\sigma_n^2),
\end{equation}
\label{eq:chi2}
where $y_n$ are the data, $\mu$ is the mean model and $\sigma_n$ are the
Gaussian uncertainites on the data.
The equivalent equation in matrix notation is
\begin{equation}
\ln \mathcal{L} = -\frac{1}{2}\bf{r}^T\bf{C}^{-1}\bf{r}-\ln|\bf{C}|
    + \mathrm{constant},
\end{equation}
\label{eq:lhf1}
where $\bf{r}$ is the vector of residuals and $\bf{C}$ is the covariance
matrix,
\begin{eqnarray}
	\mathbf{C} &=& \left (\begin{array}{cccc}
	\sigma^2_1 & \sigma_{2, 1} & \cdots & \sigma_{N, 1} \\
	\sigma_{1, 2} & \sigma^2_2 & \cdots & \sigma_{N, 2} \\
    && \vdots & \\
	\sigma_{1, N} & \sigma_{2, N} & \cdots & \sigma^2_N
\end{array}\right )
\end{eqnarray}
In the case where the uncertainties are uncorrelated, the noise is `white',
(which is a frequent assumption made by astronomers and is sometimes
justified) and the off-diagonal elements of the covariance matrix are zero.
However, in the case where there is evidence for correlated
`noise'\footnote{In our case the `noise' is actually the model!}, as in the
case of \Kepler\ light curves, those off-diagonal elements are non-zero.
With GP regression, a covariance matrix generated by the kernel function,
${\bf K}$ replaces ${\bf C}$ in the above equation.
Incidentally, this approach is the reverse of the regression techniques
usually employed by astronomers.
In most problems in astronomy one tries to infer the parameters that describe
the mean model and, if correlated noise is present, to marginalise over that
noise.
Here, the parameters describing the correlated noise are what we are
interested in and our mean model is simply a straight line at $y=0$.

@@

We address the difficulty of computational speed in two distinct ways: subsampling
the data and splitting the light curve into independent sections.
To subsample, we randomly select 1/30th of the points in the full light curve (an
average of $\sim$1.5 points per day).  This decreases the likelihood evaluation
time by a factor of about 50, down to about 100\,ms.
We then split the light curve into equal-sized
chunks containing approximately 300 points per section (corresponding to about
200 days), and evaluate the log-likelihood as the sum of the log-likelihoods
of the individual sections (all using the same parameters $\theta$).  This reduces
computation time because the section-based likelihood evaluation scales as $mn\ln(n)^2$, where $n$ is the number of data points per section and $m$ is the number of sections.
This method further reduces computation time for a typical light curve (subsampled
by a factor of 30) by about a factor of two, to about 50\,ms.

In order to manage the great flexibilty of the GP model and focus on
reliably retrieving the period parameter, we also impose priors
on the non-period GP parameters that we
learned during the development of this method.
Beginning with very broad priors (uniform in log space between
-20 and 20) on all the parameters, we first explored what the
typical values of the non-period parameters tended to be when the
injected periods were successfully recovered.  We also experimented with
constraining the allowed ranges of the parameters, after discovering
that some regions of parameter space (such as large values of $A$ and $\Gamma$
and small values of $l$) tended to allow fits that ignored the desired
periodicity.  The priors and bounds we used for this work are listed in
table \ref{tab:priors}.  We note that we do not have any specific
quantitative justification for the use of these priors,
except that they produce good results for this particular test set of
light curves.  We thus caution that other data sets might have different
noise properties, potentially requiring modifications to these priors
on the non-period parameters.  For period, we tried two different methods:
an uninformed (log-flat) prior between 0.5\,d and 100\,d, and an ACF-informed
prior (details in Appendix.)
% We note that our default prior for period is
% log-uniform between the bounds, but that we were also able to
% further improve our results by using an ACF-inspired prior.

@@

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{figures/compare_mcmc.pdf}
\includegraphics[width=6in, clip=true]{figures/comparison_emcee3_17_1.png}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique.
Since the posterior PDFs of rotation periods are often non-Gaussian,
the points plotted here are the highest likelihood samples.
The uncertainties are the 16th and 84th percentiles.}
\label{fig:compare_mcmc}
\end{center}
\end{figure*}

>>>>

In order to recover the rotation period of a simulated light curve using
Gaussian processes, we sample the following posterior PDF:
\begin{equation}
p({\bm \theta}\,|\,y) \propto \mathcal L(y\,|\,{\bm \theta}) p({\bm \theta}),
\end{equation}
\label{eq:posterior}
where $y$ are the light curve data, $\bm \theta$ are the hyperparameters
of the kernel described in equation \ref{eq:QP}, $\mathcal L$ is the
GP likelihood function, and $p({\bm \theta})$ is the prior on the
hyperparameters.  The GP likelihood is similar to the simple Gaussian likelihood
function used for optimisation problems where the uncertainties are
Gaussian and uncorrelated. The latter can be written
\begin{equation}
\ln \mathcal{L} = -\frac{1}{2}\sum_{n=1}^N\frac{(y_n-\mu)^2}{\sigma_n^2}
    - \frac{N}{2}\ln(2\pi\sigma_n^2),
\end{equation}
\label{eq:chi2}
where $y_n$ are the data, $\mu$ is the mean model and $\sigma_n$ are the
Gaussian uncertainites on the data.
The equivalent equation in matrix notation is
\begin{equation}
\ln \mathcal{L} = -\frac{1}{2}\bf{r}^T\bf{C}^{-1}\bf{r}-\ln|\bf{C}|
    + \mathrm{constant},
\end{equation}
\label{eq:lhf1}
where $\bf{r}$ is the vector of residuals and $\bf{C}$ is the covariance
matrix,
\begin{eqnarray}
	\mathbf{C} &=& \left (\begin{array}{cccc}
	\sigma^2_1 & \sigma_{2, 1} & \cdots & \sigma_{N, 1} \\
	\sigma_{1, 2} & \sigma^2_2 & \cdots & \sigma_{N, 2} \\
    && \vdots & \\
	\sigma_{1, N} & \sigma_{2, N} & \cdots & \sigma^2_N
\end{array}\right )
\end{eqnarray}
In the case where the uncertainties are uncorrelated, the noise is `white',
(which is a frequent assumption made by astronomers and is sometimes
justified) and the off-diagonal elements of the covariance matrix are zero.
However, in the case where there is evidence for correlated
`noise'\footnote{In our case the `noise' is actually the model!  Incidentally, this approach is the reverse of the regression techniques
usually employed by astronomers.
In most problems in astronomy one tries to infer the parameters that describe
the mean model and, if correlated noise is present, to marginalise over that
noise.
Here, the parameters describing the correlated noise are what we are
interested in and our mean model is simply a straight line at $y=0$.}, as in the
case of \Kepler\ light curves, those off-diagonal elements are non-zero.
With GP regression, a covariance matrix generated by the kernel function,
${\bf K}$ replaces ${\bf C}$ in the above equation.

@@

We address the difficulty of computational speed in two distinct ways: subsampling
the data and splitting the light curve into independent sections.
To subsample, we randomly select 1/30th of the points in the full light curve (an
average of $\sim$1.5 points per day).  This decreases the likelihood evaluation
time by a factor of about 50, down to about 100\,ms.
We then split the light curve into equal-sized
chunks containing approximately 300 points per section (corresponding to about
200 days), and evaluate the log-likelihood as the sum of the log-likelihoods
of the individual sections (all using the same parameters ${\bm \theta}$).  This reduces
computation time because the section-based likelihood evaluation scales as $mn\ln(n)^2$, where $n$ is the number of data points per section and $m$ is the number of sections.
This method further reduces computation time for a typical light curve (subsampled
by a factor of 30) by about a factor of two, to about 50\,ms.

In order to manage the great flexibilty of the GP model and focus on
reliably retrieving the period parameter, we also impose priors
on the non-period GP parameters that we
learned during the development of this method.
Beginning with very broad priors (uniform in log space between
-20 and 20) on all the parameters, we first explored what the
typical values of the non-period parameters tended to be when the
injected periods were successfully recovered.  We also experimented with
constraining the allowed ranges of the parameters, after discovering
that some regions of parameter space (such as large values of $A$ and $\Gamma$
and small values of $l$) tended to allow fits that ignored the desired
periodicity.  The priors and bounds we used for this work are listed in
table \ref{tab:priors}.  We note that we do not have any specific
quantitative justification for the use of these priors,
except that they produce good results for this particular test set of
light curves.  We thus caution that other data sets might have different
noise properties, potentially requiring modifications to these priors
on the non-period parameters (e.g., see Section \ref{sec:kepler}).  
For period, we tried two different methods:
an uninformed (log-flat) prior between 0.5\,d and 100\,d, and an ACF-informed
prior (details in Appendix.)
% We note that our default prior for period is
% log-uniform between the bounds, but that we were also able to
% further improve our results by using an ACF-inspired prior.

@@

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{figures/compare_mcmc.pdf}
\includegraphics[width=6in, clip=true]{figures/comparison_emcee3_24_0.png}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique.
Since the posterior PDFs of rotation periods are often non-Gaussian,
the points plotted here are the highest likelihood samples.
The uncertainties are the 16th and 84th percentiles.}
\label{fig:compare_mcmc}
\end{center}
\end{figure*}
